---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
#Summary

This is the markdown document for Reproducible Research Course at Coursera for participant Sami Ahma-aho.
Github:
https://github.com/samfield/RepData_PeerAssessment1.git


## Loading and preprocessing the data

I will use some basic packages for plotting and data handling, the following loads those packages:
```{r load, echo=TRUE}
require(data.table)
require(dplyr)
require(ggplot2)
require(lattice)
```

It is assumed that the working directory of the user contains the same package of data as in the original repository, namely the activity.zip.

First we will load the data. By using the unzip command

```{r unzip,echo=TRUE}
#directories;
#define the where the zip file is downloaded
temp <- paste(getwd(), "activity.zip", sep = "/")
#unzip the loaded file to the working directory then load a data.table with fread function
dt <- fread(unzip(temp), verbose = TRUE)
#turn date as a date variable
dt[,date:=as.Date(date, "%Y-%m-%d")]
#turn steps to numeric
dt[,steps:=as.numeric(steps)]
#the summary for our data is 
summary(dt)
```

## What is mean total number of steps taken per day?

__Sum of steps per day__

```{r total_nbr_of_steps_per_day, echo=TRUE}
#calculation of sum of steps per day, removing na values, but reporting whole days with no values
dt[,.(sum_steps=sum(steps, na.rm = TRUE)), by=date]
```

__Total number of steps daily__

```{r histogram_daily, echo=TRUE}
#plotting total number of steps daily
#the data is a steps by dates
ggplot(data=dt[is.na(steps)==FALSE,.(sum_of_steps=sum(steps, na.rm = TRUE)), by=date], aes(sum_of_steps)) + geom_histogram() +xlab("Number of steps") +ylab("Frequency of days") +ggtitle("Histogram of sum of steps by day")
```

__Mean and median steps for total steps per day__

```{r mean_and_median_steps_per_day, echo=TRUE}
#calculation of mean of steps for total sum of steps per day
dt[,.(sum_steps=sum(steps, na.rm = TRUE)), by=date][,.(steps_mean=mean(sum_steps, na.rm = TRUE))]
#calculation of median of steps per day
dt[,.(sum_steps=sum(steps, na.rm = TRUE)), by=date][,.(steps_median=median(sum_steps, na.rm = TRUE))]
```


## What is the average daily activity pattern?

__Time series of average daily activity__

```{r average_plot, echo=TRUE}
#make the plot of average daily activity
ggplot(data=dt[is.na(steps)==FALSE,.(average=mean(steps, na.rm = TRUE)), by=.(interval)],aes(x=interval,y=average)) + geom_line() +ggtitle("Average daily pattern")
```


__Maximum steps averaged over days(of interval)__

```{r max_steps, echo=TRUE}
dt[is.na(steps)==FALSE,.(average=mean(steps, na.rm = TRUE)),by=.(interval)][,max(average)]
```


__The interval of maximum amount of steps__

```{r most_steps,echo=TRUE}
dt[is.na(steps)==FALSE,.(average=mean(steps, na.rm = TRUE)),by=.(interval)][max(average),interval]
```


## Imputing missing values

###Reporting missing values per steps,date and interval 

```{r totat_amount_na, echo=TRUE}
#missing values for steps
dt[is.na(steps)==TRUE,.N]
#missing values for date
dt[is.na(date)==TRUE,.N]
#missing values for interval
dt[is.na(interval)==TRUE,.N]
```

Total of `r dt[is.na(steps)==TRUE,.N]` missing values.

Missing values are imputed with the mean of the series.

```{r impute_mean,echo=TRUE}
#create a new dataset
#impute mean steps by mean of day
dt_impute <- dt
mean_steps <- mean(dt$steps, na.rm = TRUE)
dt_impute[is.na(steps)==TRUE,steps:=mean_steps]
```

__Histogram for the imputed series__

```{r new_hist,echo=TRUE}
ggplot(data=dt_impute[is.na(steps)==FALSE,.(sum_of_steps=sum(steps, na.rm = TRUE)), by=date], aes(sum_of_steps)) + geom_histogram() +xlab("Number of steps") +ylab("Frequency of days") +ggtitle("Histogram of sum of steps by day, imputed dataset")
```

__Comparison between original and imputed datasets - Mean and median steps for total steps per day__

```{r mean_and_median_steps_per_day_imputed, echo=TRUE}
#calculation of mean of steps for total sum of steps per day
dt_impute[,.(sum_steps=sum(steps, na.rm = TRUE)), by=date][,.(steps_mean=mean(sum_steps, na.rm = TRUE))]
#calculation of median of steps per day
dt_impute[,.(sum_steps=sum(steps, na.rm = TRUE)), by=date][,.(steps_median=median(sum_steps, na.rm = TRUE))]
```

The values of mean and median for the imputed series differ from the original series. The median and mean are the same after imputation.

## Are there differences in activity patterns between weekdays and weekends?

```{r calc_date, echo=TRUE}
#lets use the data.table function for weekdays 
#calculate what day of the week the obs is
dt_impute[,weekday:=as.POSIXlt(date)$wday +1L]
#check
dt_impute
#turn as a factor variable
dt_impute[weekday <6, weekend:="weekday"]
dt_impute[weekday >5, weekend:="weekend"]
dt_impute[,weekend:=as.factor(weekend)]
```

```{r weekend_plot, echo=TRUE, fig.width=11}
#the plot requested
ggplot(data=dt_impute[,.(average=mean(steps, na.rm = TRUE)), by=.(interval,weekend)],aes(x=interval,y=average)) + geom_line() +ggtitle("Average daily pattern") + facet_grid(weekend~.)
#a better plot for comparing differences
ggplot() + geom_line(data=dt_impute[,.(average=mean(steps, na.rm = TRUE)), by=.(interval,weekend)],aes(x=interval,y=average, color=weekend)) +ggtitle("Average daily pattern")
```

